\documentclass[notitlepage]{report}
\usepackage {listings}
\usepackage{ulem}
\renewcommand{\thesection}{\arabic{part}.\arabic{section}}
\makeatletter
\@addtoreset{section}{part}
\makeatother 

\begin{document}
\title{Emulation Implementation Notes}
%\author{Charles Anthony and others}
%\date{2015-08-21}
\maketitle

% \part{Source Code Layout}
%   \section{dps8\_sys}
%     \subsection{dps8\_cable}
%   \section{dps8\_cpu}
%     \subsection{dsp8\_iefp}
%     \subsection{dps8\_addrmods}
%     \subsection{dps8\_append}
%     \subsection{dps8\_bar}
%     \subsection{dps8\_ins}
%     \subsection{dps8\_eis}
%      \subsection{dps8\_math}
%     \subsection{dps8\_opcodetable}
%     \subsection{dps8\_faults}
%     \subsection{dps8\_decimal}
%   \section{dps8\_scu}
%   \section{dps8\_iom}
%     \subsection{dps8\_console}
%     \subsection{dps8\_disk}
%     \subsection{dps8\_mt}
%     \subsection{dps8\_fnp, fnp\_ipc}
%     \subsection{dps8\_lp}
%     \subsection{dps8\_crdrdr}
%   \section{dps8\_utils}
%     \subsection{shm}
%   \section{Other}

% \part{CPU operation}
%   \section{simh}
%   \section{CPU emulation organization}
%   \section{FETCH\_cycle}
%   \section{EXEC\_cycle}
%   \section{Instruction execution: `executeInstruction'}
%   \section{Instruction execution: `doInstruction'}
%     \subsection{LCA Load Complement A}
%     \subsection{LREG Load Registers}
%     \subsection{STA Store A}
%     \subsection{STXn Store Index Register n}
%     \subsection {TRA Transfer}
%   \section{RPT/RPD}
%   \section{XEC/XED}
%   \section{EIS}
%   \section{Computed Address Formation}
%   \section{Append Unit}
%   \section{Interrupts}
%   \section{Faults}
%     \subsection{doFault()}
%     \subsection{RCU}
%   \section{CU implementation}

% \part{"As-built"}
%   \section{Variations between the hardware and the emulator}

\part{Source Code Layout}

\section{dps8\_sys} 

This module handles the abstract "entire system." The bulk of the code is 
initialization and simh command processing hooks.

\subsection{dps8\_cable}

Hardware device cabling emulation

\section{dps8\_cpu} 

The CPU emulator

\subsection{dsp8\_iefp}

CPU Memory access

\subsection{dps8\_addrmods}

CPU Address Modifcation

\subsection{dps8\_append}

CPU Append Unit

\subsection{dps8\_bar}

CPU BAR address computation

\subsection{dps8\_ins}

CPU instructions

\subsection{dps8\_eis}

CPU EIS instructions

\subsection{dps8\_math}

CPU math support routines

\subsection{dps8\_opcodetable}

CPU opcode table; drives address preparation and mode checking

\subsection{dps8\_faults}

CPU fault handling code

\subsection{dps8\_decimal}

CPU Decimal math support routines

\section{dps8\_scu}

The SCU emulator

\section{dps8\_iom}

The IOM emulator

\subsection{dps8\_console}

Operator console device

\subsection{dps8\_disk}

Disk device

\subsection{dps8\_mt}

Tape device

\subsection{dps8\_fnp, fnp\_ipc}

FNP device

\subsection{dps8\_lp}

Printer device

\subsection{dps8\_crdrdr}

Card reader device

\section{dps8\_utils}

Common utility routines

\subsection{shm}

Host shared memory management

\section{Other}

dps8\_utils Common support code \\
dps8\_clk  Unused simh clock hooks \\
dps8\_loader Segment loader for unit tests \\
dps8\_stddev Vestigial \\
dps8\_fxe Faux Multics Execution module

\part{CPU operation}

\section{simh}

When simh is running the CPU it calls sim\_instr(), which is the CPU emulator entry point.

sim\_instr() loops, executing emulated instructions until some emulation halting condition is met, or some simh component signals for a pause due to an external
event.

\section{CPU emulation organization}

sim\_instr first establishes a setjmp context; this is used primarily by RCU 
mechanism to restart instruction processing after restoring a saved system
state due to a fault or an interrupt.

The cpu emulation is written as a state machine. The longjmp parameter is 
used to setup the desired state; at initial entry, setjmp returns a zero, 
and the appropriate setup is done.

\begin{verbatim}
#define JMP\_ENTRY       0

    int val = setjmp(jmpMain);

    switch (val)
      {
        case JMP\_ENTRY:
        case JMP\_REENTRY:
            reason = 0;
            break;
        case JMP\_NEXT:
            goto nextInstruction;
        case JMP\_STOP:
            reason = STOP\_HALT;
            goto leave;
        case JMP\_SYNC\_FAULT\_RETURN:
            goto syncFaultReturn;
        case JMP\_REFETCH:
            cpu . wasXfer = false;
            setCpuCycle (FETCH\_cycle);
            break;
        case JMP\_RESTART:
            setCpuCycle (EXEC\_cycle);
            break;
        default:
          sim\_printf ("longjmp value of %d unhandled\n", val);
          goto leave;
      }

\end{verbatim}

The cpu emulation then enters a "do {...} while (reason == 0)" loop,
which cycles the CPU through its states.

The top of the loop checks the simh components for events that need
to be handled (simh\_hooks()), polls various subsystems for service requests
(emulator console commands to be processed, incoming FNP messages and 
operator console input).

\begin{verbatim}
        reason = 0;

        // Process deferred events and breakpoints
        reason = simh\_hooks ();
        if (reason)
          {
            //sim\_printf ("reason: %d\n", reason);
            break;
          }

        static uint queueSubsample = 0;
        if (queueSubsample ++ > 10240) // ~ 100Hz
          {
            queueSubsample = 0;
            scpProcessEvent ();
            fnpProcessEvent ();
            consoleProcess ();
          }
        if (check\_attn\_key ())
          console\_attn (NULL);
\end{verbatim}

It then checks for Timer Register runout, setting the group 7 fault flag
if needed. (Group 7 faults are distinguished as not resulting from
instruction execution, but external events; and has such are handled
synchronously between instruction execution steps, rather then 
interrupting mid-instruction. 

\begin{verbatim}
    bool overrun;
    UNUSED word27 rTR = getTR (& overrun);
    if (overrun)
      {
        ackTR ();
        if (switches . tro\_enable)
          setG7fault (FAULT\_TRO, 0);
      }
\end{verbatim}

Next, it then checks for lockup (the operating system has not enabled 
interrupts for more then 32 ms.), and faults if needed.

\begin{verbatim}
        lufCounter ++;
        // Assume CPU clock ~ 1Mhz. lockup time is 32 ms
        if (lufCounter > 32000)
          {
            lufCounter = 0;
            doFault (FAULT\_LUF, 0, "instruction cycle lockup");
          }
\end{verbatim}

Lastly, it checks the CPU state and branches to the appropriate code.

\begin{verbatim}
    switch (cpu . cycle)
      {
        case INTERRUPT\_cycle:
          ....
      }
\end{verbatim}

The states are:

\begin{description}
\item [FETCH\_cycle] Fetch the next instruction
\item [EXEC\_cycle] Execute an instruction
\item [INTERRUPT\_cycle] Fetch an Interrupt instruction pair
\item [INTERRUPT\_EXEC\_cycle] Execute the even instruction of an interrupt pair
\item [INTERRUPT\_EXEC2\_cycle] Execute the odd instruction of an interrupt pair
\item [FAULT\_cycle]Fetch an Fault instruction pair
\item [FAULT\_EXEC\_cycle] Execute the even instruction of a fault pair
\item [FAULT\_EXEC2\_cycle] Execute the odd instruction of a fault pair
\end{description}

The normal instruction flow is alternating FETCH and EXEC cycles.

\section{FETCH\_cycle}

The fetch cycle first checks for pending interrupts and group 7 faults, 
according to complex eligibility rules (AL39, pg 327, "Interrupt Sampling.")

\begin{verbatim}
    if ((! cpu . wasInhibited) &&
        (PPR . IC % 2) == 0 &&
        (! cpu . wasXfer) &&
        (! (cu . xde | cu . xdo | cu . rpt | cu . rd)))
      {
        cpu . interrupt\_flag = sample\_interrupts ();
        cpu . g7\_flag = bG7Pending ();
      }
    // The cpu . wasInhibited accumulates across the even and 
    // odd intruction. If the IC is even, reset it for
    // the next pair.

    if ((PPR . IC % 2) == 0)
      cpu . wasInhibited = false;
\end{verbatim}

If a eligible interrupt is pending, the CPU state is switched to 
INTERRUPT\_cycle.

\begin{verbatim}
    if (cpu . interrupt\_flag)
      {
        setCpuCycle (INTERRUPT\_cycle);
        break;
      }
\end{verbatim} 

Likewise, if a Group 7 faults is pending, cause a fault.

\begin{verbatim}
    if (cpu . g7\_flag)
      {
        cpu . g7\_flag = false;
        doG7Fault ();
      }
\end{verbatim}

There is now code to process the XEC and XED instructions; the idea here is
that the processing of the XEC and XED instructions loads the target 
instructions into the Control Unit IWB and IODD words, and that the 
fetch cycle is a no-op, as the instructions have already been 
fetched.

If not the XEC or XED the case, the instruction is fetched into the
CU IWB word.

\begin{verbatim}
    else
      {
        processorCycle = INSTRUCTION\_FETCH;
        clr\_went\_appending ();
        fetchInstruction (PPR . IC);
      }
\end{verbatim}

Now that the instruction is in the IWB, switch to EXEC state.

\begin{verbatim}
    setCpuCycle (EXEC\_cycle);
    break;
\end{verbatim}

\section{EXEC\_cycle}

The execute cycle starts right off with:

\begin{verbatim}
    t\_stat ret = executeInstruction ();
\end{verbatim}

A return value of greater than 0 indicates that the CPU needs to halt:

\begin{verbatim}
    if (ret > 0)
      {
        reason = ret;
        break;
      }
\end{verbatim}

A return value of CONT\_TRA indicates that a transfer instruction 
was executed, which requires different handling. A transfer 
may be of interest to the handling of
transfer into append or BAR modes, so the `wasXfer' flag is set; and
the code that increments the IC is skipped. The CPU state is set to
EXEC\_cycle, and it all repeats.

\begin{verbatim}
    if (ret == CONT\_TRA)
      {
        cpu . wasXfer = true;
        setCpuCycle (FETCH\_cycle);
        break;   // don't bump PPR.IC, instruction already did it
      }
    cpu . wasXfer = false;
\end{verbatim}


Next, the IC is incremented to point to the next instruction; any
EIS operands that the instruction had are also skipped over.
The CPU state is set to FETCH\_cycle, and it all repeats.

\begin{verbatim}
    PPR.IC ++;
    if (ci->info->ndes > 0)
      PPR.IC += ci->info->ndes;

    cpu . wasXfer = false;
    setCpuCycle (FETCH\_cycle);
    break;
\end{verbatim}


\section{Instruction execution: `executeInstruction'}

Instruction execution is managed by `executeInstruction()'.

When `executeInstruction' starts, the instruction has been loaded
into the CU IWB (Control Unit Instruction Working Buffer).

First, the instruction is decoded, extracting the operation code, address
field, tag field and the A and I bits into the `currentInstruction'
structure; and setting the 'info' member to point to the operation's
entry in the opcode table.

\begin{verbatim}
    decodeInstruction(cu . IWB, ci);
\end{verbatim}

The 'info' member contains a wide variety of details about the instruction,
including:

\begin{description}
\item [PREPARE\_CA] Instruction will need the operand's computed address.
\item [READ\_OPERAND] Instruction will read the operand.
\item [WRITE\_OPERAND] Instruction will write the operand.
\item [NO\_RPT] Not allowed in a repeat instruction.
\item [PRIV\_INS] Privileged instruction.
\item [] The number of EIS operands.
\item [] Allowed tag values.
\end{description}

The flags in the `info' structure are used to check instruction restrictions, 
such as privilege and allowed modifiers; violation causes a fault.

\begin{verbatim}
    if ((ci -> info -> flags & PRIV\_INS) && ! is\_priv\_mode ())
        doFault (FAULT\_IPR, ill\_proc, 
                 "Attempted execution of privileged instruction.");
    // No CI/SC/SCR allowed
    if (ci->info->mods == NO\_CSS)
    {
        if (\_nocss[ci->tag])
            doFault(FAULT\_IPR, ill\_mod, "Illegal CI/SC/SCR modification");
    }
    // No DU/DL/CI/SC/SCR allowed
    else if (ci->info->mods == NO\_DDCSS)
    {
        if (\_noddcss[ci->tag])
            doFault(FAULT\_IPR, ill\_mod, "Illegal DU/DL/CI/SC/SCR modification");
    }
    // No DL/CI/SC/SCR allowed
    else if (ci->info->mods == NO\_DLCSS)
    {
        if (\_nodlcss[ci->tag])
            doFault(FAULT\_IPR, ill\_mod, "Illegal DL/CI/SC/SCR modification");
    }
    // No DU/DL allowed
    else if (ci->info->mods == NO\_DUDL)
    {
        if (\_nodudl[ci->tag])
            doFault(FAULT\_IPR, ill\_mod, "Illegal DU/DL modification");
    }
\end{verbatim}

Next, an assortment of initializations occurs, setting various registers
to the operand address field, and initializing the TPR register.

\begin{verbatim}
    TPR.CA = address;
    iefpFinalAddress = TPR . CA;
    rY = TPR.CA;

    TPR.TRR = PPR.PRR;
    TPR.TSR = PPR.PSR;
\end{verbatim}

If the instruction has EIS operands, they are read into holding variables.

\begin{verbatim}
    if (info -> ndes > 0)
      {
        for(int n = 0; n < info -> ndes; n += 1)
          {
// XXX This is a bit of a hack; In general the code is good about 
// setting up for bit29 or PR operations by setting up TPR, but
// assumes that the 'else' case can be ignored when it should set
// TPR to the canonical values. Here, in the case of a EIS instruction
// restart after page fault, the TPR is in an unknown state. Ultimately,
// this should not be an issue, as this folderol would be in the DU, and
// we would not be re-executing that code, but until then, set the TPR
// to the condition we know it should be in.
            TPR.TRR = PPR.PRR;
            TPR.TSR = PPR.PSR;
            Read (PPR . IC + 1 + n, & ci -> e . op [n], EIS\_OPERAND\_READ, 0);
          }
        // This must not happen on instruction restart
        if (! (cu . IR & I\_MIIF))
          {
            du . CHTALLY = 0;
            du . Z = 1;
          }
      }
\end{verbatim}

If the instruction expects the address field to be converted to the
Computed Address, do that.

\begin{verbatim}
        if (ci->info->flags & PREPARE\_CA)
          {
            doComputedAddressFormation ();
            iefpFinalAddress = TPR . CA;
          }
\end{verbatim}

Otherwise if the instruction wants the operand value, do that. `readOperands' will handle single, double, eight and sixteen word operands. The value is held in `CY', `Ypair', `Yblock8' or `Yblock16' holding registers, as appropriate. Read operands handles all aspects of indirection and address appending.

\begin{verbatim}
        else if (READOP (ci))
          {
            doComputedAddressFormation ();
            iefpFinalAddress = TPR . CA;
            readOperands ();
          }
\end{verbatim}

Now that the operands have sorted, the instruction is executed.

\begin{verbatim}
    t\_stat ret = doInstruction ();
\end{verbatim}

If an transfer instruction has the A bit set, and accesses the Append Unit
during the Computed Address formation, the processor is switch to Append
mode.

\begin{verbatim}
    if (info->ndes == 0 && a && (info->flags & TRANSFER\_INS))
      {
        if (get\_addr\_mode () == BAR\_mode)
          set\_addr\_mode(APPEND\_BAR\_mode);
        else
          set\_addr\_mode(APPEND\_mode);
      }
\end{verbatim}

Finally, it the instruction writes the operand, do that.

\begin{verbatim}
    if (WRITEOP (ci))
      {
        if (! READOP (ci))
          {
            doComputedAddressFormation ();
            iefpFinalAddress = TPR . CA;
          }
        writeOperands ();
      }
\end{verbatim}

\section{Instruction execution: `doInstruction'}

First, initialize the EIS state registers.

\begin{verbatim}
    if (i->info->ndes > 0)
    {
        i->e.ins = i;
        i->e.addr[0].e = &i->e;
        i->e.addr[1].e = &i->e;
        i->e.addr[2].e = &i->e;

        i->e.addr[0].mat = OperandRead;   // no ARs involved yet
        i->e.addr[1].mat = OperandRead;   // no ARs involved yet
        i->e.addr[2].mat = OperandRead;   // no ARs involved yet
    }
\end{verbatim}

And switch based on the opcode extension bit; `Basic' and `EIS' are
misleading here; the opcode extension bit has a minimal correlation to
the EIS instruction opcode layout, but by separating the two cases,
the code becomes a bit more readable, Both routines have the same
preamble:

\begin{verbatim}
    DCDstruct * i = & currentInstruction;
    uint opcode  = i->opcode;  // get opcode

    switch (opcode)
    {
       ....
    }
\end{verbatim}

We will look at a few of the instructions so as to understand the
general function of 'doInstruction.'

\subsection{LCA Load Complement A}

`readOperands()' has already retrieved the operand value, and left it in CY.
The utility routine `compl36' complements the value and sets the IR flags,
and the assignment operation places the complemented value in the A register.

\begin{verbatim}
    case 0335:  // lca
      rA = compl36 (CY, & cu . IR);
      break;
\end{verbatim}

\subsection{LREG Load Registers}

LREG loads the A, Q, E, and index registers from a Y-block8. Again, 
`readOperands()' has loaded the values into Yblock8.

\begin{verbatim}
    case 0073:   ///< lreg
      rX[0] = GETHI(Yblock8[0]);
      rX[1] = GETLO(Yblock8[0]);
      rX[2] = GETHI(Yblock8[1]);
      rX[3] = GETLO(Yblock8[1]);
      rX[4] = GETHI(Yblock8[2]);
      rX[5] = GETLO(Yblock8[2]);
      rX[6] = GETHI(Yblock8[3]);
      rX[7] = GETLO(Yblock8[3]);
      rA = Yblock8[4];
      rQ = Yblock8[5];
      rE = (GETHI(Yblock8[6]) >> 10) & 0377;   // need checking
      break;
\end{verbatim}

\subsection{STA Store A}

Since `writeOperands()' will do the actual writing, all STA needs to
do is copy A to CY.

\begin{verbatim}
    case 0755:  // sta
      CY = rA;
      break;
\end{verbatim}

\subsection{STXn Store Index Register n}

For many opcodes, the low bits of the opcode contains indexing 
information.

\begin{verbatim}
    case 0740:  ///< stx0
    case 0741:  ///< stx1
    case 0742:  ///< stx2
    case 0743:  ///< stx3
    case 0744:  ///< stx4
    case 0745:  ///< stx5
    case 0746:  ///< stx6
    case 0747:  ///< stx7
      {
        uint32 n = opcode & 07;  // get n
        SETHI(CY, rX[n]);
      }
    break;
\end{verbatim}

\subsection {TRA Transfer}

For the TRA instruction, the computed address is placed in the PPR, and the
function return value is used to signal that a transfer is to occur.

\begin{verbatim}
    case 0710:  ///< tra
      PPR.IC = TPR.CA;
      PPR.PSR = TPR.TSR;
      return CONT\_TRA;
\end{verbatim}

\section{RPT/RPD}

The emulated RPT and RPD instruction execution is complex.

When the RPT or RPD instruction is executed:

1. The low six bits of the instruction are copied to CU DELTA register.

2. The C flag is extracted from the instruction (bit 10); if set,
the high eighteen bits (tally, A, B, C and
termination condition flags) of the instruction are copied to the X0 register.

3. The CU RF flag (repeat first) is set.

4. RPT sets the CU RPT flag; RPD the CU RD flag.

The `doInstruction()' routine (which executed the RPD/RPT instruction returns to
`executeInstruction();' as part of it post-execution processing, it examines the CU RF 
(repeat first) flag.

If CU RF is set, then the CU RD (repeat double) and the even/odd bit of PPR.IC is examined.
If RD is not set or, if it is and this is not the even (first) instruction of the repeat pair,
then a local flag (rf) is set.

\begin{verbatim}
bool rf = cu . repeat_first;
    if (rf && cu . rd && icEven)
      rf = false;
\end{verbatim}

The idea is that `rf' will be set if we have just executed the repeated instruction, or the 
second instruction of the repeat double instruction for the first time.

Next, if `rf' is not set, the index register(s) must be updated and termination
conditions are checked:

a. If the CU RPT flag is on, then the index register referenced by the just executed instruction 
is incremented by CU DELTA.

b. If the CU RPD flag is on the index registers are not incremented until both of the 
repeated instructions have been executed; this is determined by the IC being odd.

c. If the CU RPD flag is on and the IC is odd and the B flag is set in X0, the index
register referenced by the second instruction is incremented by CU DELTA 

d. If the CU RPT flag is on or if the CU RPD flag is on and the IC is odd, then the 
termination flags in X0 are examined and indicated tests are made (tally runout, not
tally runout, zero, not zero, negative, not negative, carry out, not carry out and
overflow). If any of the termination conditions are met, the CU RPT and CU RD flags
are cleared.

`fetchInstruction()' checks the CU RPT, CU RD, CU RF and IC odd bit flags:

If CU RF is set and CU RD is set and the IC is odd, the instruction is fetched into
CU IRODD register.

a. If CU RF is set and CU RPT is set, the instruction is fetched into the IWB.

b. If CU RF is not set, and either CU RPT is set or CU RD is set and IC is odd, no instruction is fetched.

When the repeated instructions are by `executeInstruction()', it checks the CU RPT, CU RD and IC odd flags and executes either the instruction in IWB or in IRODD, as indicated.

`executeInstruction()' also checks the repeated instruction's tag field to enforce illegal
modifier faults.

Next, `executeInstruction()' must handle the 'repeat first' setup. If CU RPT or CU RD are set, and
CU RF is set:

a. If CU RPT is set, clear CU RF.

b. If CU RD is set and the IC is odd, clear CU RF.

c. If CU RPT is set or CU RD is set, offset in the instruction is added to the referenced X register

Lastly, the EXEC\_cycle code checks the CU flags and if the instruction is to be repeated,
adjusts the IC as needed and starts a fetch cycle.

The overall flow is:

\begin{verbatim}
RPT:

   RPT is executed
     Set CU DELTA
     Set X0
     Set CU RF
     Set RPT

   Post instruction execution
     CU RF is set

   EXEC_cycle post-execution
     CU RF is set

   Fetch next instruction
     CU RPT and CU RF is set; fetch the instruction into IWB

   Execute instruction
     CU RPT set; execute instruction in IWB
     Check repeated instruction tag field
     CU RF is set
       Clear CU RF
       Add instruction offset field to referenced X register
     The instruction to be repeated is executed

A:
   Post instruction execution
     CU RF is not set and CU RPT is set
       Increment the referenced X register by CU delta
       Check termination flags and conditions
         If the repeat is be terminated, clear CU RPT

   EXEC_cycle post-execution
     CU RF is not set and CU RPT is
       Switch to FETCH_cycle to repeat the instruction

   Fetch next instruction
     CU RPT is set and CU RF is not set; the instruction is
     already in IWB.

   Execute instruction
     CU RPT set; execute instruction in IWB
     Check repeated instruction tag field
     CU RF is not set
     The instruction to be repeated is executed

   Go to A:


RPD:

   RPD is executed
     Set CU DELTA
     Set X0
     Set CU RF
     Set RPD

; repeat first is set

   Post instruction execution
     CU RF is set

   EXEC_cycle post-execution
     CU RF is set

; first repeated instruction, first time

   Fetch next (first) instruction
     CU RPD and CU RF is set
       IC is even
         fetch the instruction into IWB

   Execute (first) instruction
     CU RPD is set and the IC is even; execute instruction in IWB
     Check repeated instruction tag field
     CU RF is set, but IC is even; don't clear CU RF
     CU RF is set
       Add instruction offset field to referenced X register
     The first instruction to be repeated is executed

   Post instruction execution
     CU RF is set, CU RPD is set, and the IC is even

   EXEC_cycle post-execution
     CU RF is set

; second repeated instruction, first time

   Fetch next (second) instruction
     CU RPD is set and CU RF is set and IC is odd
       Fetch second instuction into CU IRODD

   Execute (second) instruction
     CU RPD is set and the IC is odd; execute instruction in IRODD
     Check repeated instruction tag field
     CU RF is set, IC is odd
       clear CU RF
       Add instruction offset field to referenced X register
     The second instruction to be repeated is executed

; repeat first is not set

B:
   Post (second) instruction execution
     CU RF is not set, CU RPD is set, and the IC is odd
       Increment the referenced X register by CU delta
       Check termination flags and conditions
         If the repeat is be terminated, clear CU RPT

   EXEC_cycle post-execution
     CU RF is not set and CU RD is set
       The IC is odd; back it up to the first instruction
       Switch to FETCH_cycle to execute the first instruction

; first repeated instruction, not first time

   Fetch next (first) instruction
     CU RPD is set and CU RF is not set; instruction is in IRODD

   Execute (first) instruction
     CU RPD is set and the IC is even; execute instruction in IWB
     Check repeated instruction tag field
     CU RF is not set
     The first instruction to be repeated is executed

   Post instruction execution
     CU RF is not set, CU RPD is set, and the IC is even

   EXEC_cycle post-execution
     CU RF is net set, CU RPD is set, IC is even.
       Advance the IC to the second instruction
       Switch to FETCH_cycle to execute the second instruction
      
; second  repeated instruction, not first time

   Fetch next (second) instruction
     CU RPD is set and CU RF is set and IC is odd
       Fetch second instuction into CU IRODD

   Execute (second) instruction
     CU RPD is set and the IC is odd; execute instruction in IRODD
     Check repeated instruction tag field
     CU RF is set, IC is odd
       clear CU RF
       Add instruction offset field to referenced X register
     The second instruction to be repeated is executed

   Post (second) instruction execution
     CU RF is not set, CU RPD is set, and the IC is odd
       Increment the referenced X register by CU delta
       Check termination flags and conditions
         If the repeat is be terminated, clear CU RPT

   EXEC_cycle post-execution
     CU RF is not set
       The IC is odd; back it up to the first instruction
       Switch to EXEC_cycle to execute the first instruction

   Fetch next (first) instruction
     CU RPD is set and CU is not set; instruction already fetched

   Execute (first) instruction
     CU RPD is set and the IC is even; execute instruction in IWB
     Check repeated instruction tag field
     CU RF is not set
     The first instruction to be repeated is executed

   Go to B:
\end{verbatim}


\section{XEC/XED}

TODO

\section{EIS}

TODO

\section{Computed Address Formation}

TODO

\section{Append Unit}

TODO

\section{Interrupts}

Interrupts are handled at the start of the CPU fetch cycle:

\begin{verbatim}
    if (cpu . interrupt\_flag)
      {
        setCpuCycle (INTERRUPT\_cycle);
        break;
      }
\end{verbatim}

The DPS8M interrupt handling logic is to save the system state, place the processor in "temporary absolute 
mode", fetch a pair of instructions from the indicated location in memory and execute them. If one
of the pair of instructions is a transfer instruction, the processor is set to absolute mode. If
neither of the instructions transfers, the Control Unit is restored from the saved state and the
processor is switched back to the fetch cycle.

The INTERRUPT\_cycle handler saves the Control Unit state so the processor can return to the
state that was extant at the time the interrupt was handled. The interrupt number being serviced
is stored in the Control Unit, where the guest operating system can inspect it.

\begin{verbatim}
  // In the INTERRUPT CYCLE, the processor safe-stores
  // the Control Unit Data (see Section 3) into 
  // program-invisible holding registers in preparation 
  // for a Store Control Unit (scu) instruction, enters 
  // temporary absolute mode, and forces the current 
  // ring of execution C(PPR.PRR) to
  // 0. It then issues an XEC system controller command 
  // to the system controller on the highest priority 
  // port for which there is a bit set in the interrupt 
  // present register.  

  uint intr\_pair\_addr = get\_highest\_intr ();
  cu . FI\_ADDR = intr\_pair\_addr / 2;
  cu\_safe\_store ();
\end{verbatim}

Next, the processor is placed in  "temporary absolute mode":

\begin{verbatim}
    // Temporary absolute mode
    set\_TEMPORARY\_ABSOLUTE\_mode ();

    // Set to ring 0
    PPR . PRR = 0;
    TPR . TRR = 0;
\end{verbatim}

The interrupt pair is fetched and scheduled for execution:

\begin{verbatim}
    // get interrupt pair
    core\_read2 (intr\_pair\_addr, instr\_buf, instr\_buf + 1, \_\_func\_\_);

    cpu . interrupt\_flag = false;
    setCpuCycle (INTERRUPT\_EXEC\_cycle);
    break;
\end{verbatim}

The INTERRUPT\_EXEC\_cycle handler recovers an instruction from holding and executes it.

\begin{verbatim}
    case INTERRUPT\_EXEC\_cycle:
    case INTERRUPT\_EXEC2\_cycle:
      {
        if (cpu . cycle == INTERRUPT\_EXEC\_cycle)
          cu . IWB = instr\_buf [0];
        else
          cu . IWB = instr\_buf [1];

        t\_stat ret = executeInstruction ();
\end{verbatim}

If the instruction was a transfer instruction, set the processor to absolute mode and start
normal fetch/execute processing.

\begin{verbatim}
    if (ret == CONT\_TRA)
      {
        cpu . wasXfer = true;
        setCpuCycle (FETCH\_cycle);
        set\_addr\_mode (ABSOLUTE\_mode);
        break;
      }
\end{verbatim}

Otherwise, if the instruction just executed was the first of the pair, schedule the execution of
the second.

\begin{verbatim}
    if (cpu . cycle == INTERRUPT\_EXEC\_cycle)
      {
        setCpuCycle (INTERRUPT\_EXEC2\_cycle);
        break;
      }
\end{verbatim}

The only possibility now is that both instructions have been executed and neither transferred, so restore
the saved state and resume processing.

\begin{verbatim}
    clear\_TEMPORARY\_ABSOLUTE\_mode ();
    cu\_safe\_restore ();
    cpu . wasXfer = false;
    setCpuCycle (FETCH\_cycle);
    break;
\end{verbatim}
 
\section{Faults}

Faults fall (mostly) into two categories, ones that are generated by instruction execution (page faults,
overflow, etc.), and those that are independent of the instruction (timer run-out, lockup fault, etc.).


The Group 7 faults are handled similarly to interrupts; at the start of the CPU fetch cycle the
Group 7 fault pending flag is queried.

\begin{verbatim}
    if (cpu . g7\_flag)
      {
        cpu . g7\_flag = false;
        doG7Fault ();
      }
\end{verbatim}

`doG7Fault' is invoked rather then switching 
cycles directly.

\begin{verbatim}
void doG7Fault (void)
  {
     if (g7Faults & (1u << FAULT\_TRO))
       {
         g7Faults &= ~(1u << FAULT_TRO);
         doFault (FAULT\_TRO, 0, "Timer runout");
       }

     if (g7Faults & (1u << FAULT\_CON))
       {
         g7Faults &= ~(1u << FAULT\_CON);
         cu . CNCHN = g7SubFaults [FAULT\_CON] & MASK3;
         doFault (FAULT\_CON, g7SubFaults [FAULT\_CON], "Connect");
       }

     // Strictly speaking EXF isn't a G7 fault, put if we treat is as one,
     // we are allowing the current instruction to complete, simplifying
     // implementation
     if (g7Faults & (1u << FAULT\_EXF))
       {
         g7Faults &= ~(1u << FAULT\_EXF);
         doFault (FAULT\_EXF, 0, "Execute fault");
       }
     doFault (FAULT\_TRB, (\_fault\_subtype) g7Faults, "Dazed and confused in doG7Fault");
  }
\end{verbatim}

Both the interrupt and fault paths involve saving the Control Unit, but the fault case was abstracted into
`doFault', due to the large number of code paths leading there; while there is only the single entry into
the INTERRUPT\_cycle, and the Control Unit save is done in the INTERRUPT\_cycle state handler.

TODO

\subsection{doFault()}

TODO

\subsection{RCU}

TODO

\section{CU implementation}

\begin{tabular}{lrlll}

Word & Bits  & Name      & Description & Implementation \\
0    & 0-2   & PPR       & PPR.PRR                             & Canonical \\
     & 3-17  & PSR       & PPR.PSR                             & Canonical \\
     & 18    & P         & PPR.P                               & Canonical \\
     & 19    & XSF       & External segment flag               & Set to 0 \\
     & 20    & SDWAMM    & Match on SDWAM                      & Set to 0 (a) \\
     & 21    & SD-ON     & SDWAM enabled                       & Canonical (a) \\
     & 22    & PTWAMM    & Match on PTWAM                      & Set to 0 (a) \\
     & 23    & PT-ON     & PTWAM enabled                       & Canonical (a) \\
     & 24    & PI-AP     & Instruction fetch append cycle      & Unused \\
     & 25    & DSPTW     & Fetch descriptor segment PTW        & Canonical \\
     & 26    & SDWNP     & Fetch SDW - nonpaged                & Canonical \\
     & 27    & SDWP      & Fetch SDW - paged                   & Canonical \\
     & 28    & PTW       & Fetch PTW                           & Canonical \\
     & 29    & PTW2      & Fetch prepage PTW                   & Canonical \\
     & 20    & FAP       & Fetch final address - paged         & Canonical \\
     & 31    & FANP      & Fetch final address - nonpaged      & Canonical \\
     & 32    & FABS      & Fetch final address - absolute      & Canonical \\
     & 33-35 & FCT       & Fault counter - counts retries      & Unused \\
1    & 0     & IRO       & ACV illegal ring order              & Canonical \\
     &       & ISN       & SF illegal segment number           & Canonical \\
     & 1     & OEB       & ACV out of execute bracket          & Canonical \\
     &       & IOC       & IPR illegal op code                 & Canonical \\
     & 2     & E-OFF     & ACV execute bit is OFF              & Canonical \\
     &       & IA+IM     & IPR illegal address or modifier     & Canonical \\
     & 3     & ORB       & ACV out of read bracket             & Canonical \\
     &       & ISP       & IPR illegal slave procedure         & Canonical \\
     & 4     & R-OFF     & ACV read bit is OFF                 & Canonical \\
     &       & IPR       & IPR illegal EIS digit               & Canonical \\
     & 5     & OWB       & ACV out of write bracket            & Canonical \\
     &       & NEA       & SF nonexistent address              & Canonical \\
     & 6     & W-OFF     & ACV write bit is OFF                & Canonical \\
     &       & OOB       & SF BAR out of bounds                & Canonical \\
     & 7     & NO GA     & ACV not a gate                      & Canonical \\
     & 8     & OCB       & ACV out of call bracket             & Canonical \\
     & 9     & OCALL     & ACV outward call                    & Canonical \\
     & 10    & BOC       & ACV bad outward call                & Canonical \\
     & 11    & PTWAM\_ER & ACV PTWAM error                     & Set to 0 (a) \\
     & 12    & CRT       & ACV cross ring transfer             & Canonical \\
     & 13    & RALR      & ACV ring alarm                      & Canonical \\
     & 14    & SDWAM\_ER & ACV SDWAM error                     & Set to 0 (a) \\
     & 15    & OOSB      & ACV out of segment bounds           & Canonical \\
     & 16    & PARU      & Parity error upper                  & Set to 0 (b) \\
     & 17    & PARL      & Parity error lower                  & Set to 0 (b) \\
     & 18    & ONC1      & Operation not complete 1            & Set to 0 (c) \\
     & 19    & ONC2      & Operation not complete 2            & Set to 0 (c) \\
     & 20-23 & IA        & SC illegal action lines             & Canonical (d) \\
     & 24-26 & IACHN     & Illegal action processor port       & Set to 0. \\
     & 27-29 & CNCHN     & Connect fault processor port        & Canonical \\
     & 30-34 & F/I ADDR  & F/I vector address                  & Canonical \\
     & 35    & F/I       & F/I flag                            & Canonical \\
2    & 0-2   & TRR       & TPR.TRR                             & Canonical \\
     & 3-17  & TSR       & TPR.TSR                             & Canonical \\
     & 18-21 & PTW       & PTW enabled/match bits              & Set to 0 (a) \\
     & 22-25 & SDW       & SDW enabled/match bits              & Set to 0 (a) \\
     & 26    & 0         & 0                                   & Canonical \\
     & 27-29 & CPU       & CPU number                          & Canonical \\
     & 30-35 & DELTA     & RPT/RPD address increment           & Canonical \\
3    & 0-17  & 0         & 0                                   & Canonical \\
     & 18-21 & TSNA      & Operand 1 PR number/valid           & Set to 0 (e) \\
     & 22-25 & TSNB      & Operand 2 PR number/valid           & Set to 0 (e) \\
     & 26-29 & TSNC      & Operand 3 PR number/valid           & Set to 0 (e) \\
     & 30-35 & TEMP BIT  & TPR.TBR                             & Canonical \\
4    & 0-17  & IC        & PPR.IC                              & Canonical \\
     & 18-35 & IR        & IR (low 4 bits 0)                   & Canonical \\
5    & 0-17  & CA        & TPR.CA                              & Canonical \\
     & 18    & RF        & Repeat first cycle                  & Canonical \\
     & 19    & RPT       & Execute RPT                         & Canonical \\
     & 20    & RD        & Execute RPT                         & Canonical \\
     & 21    & RPL       & Execute RPL                         & Canonical \\
     & 22    & POT       & Prepare operand tally               & Canonical (f) \\
     & 23    & PON       & Prepare operand no tally            & Set to 0 \\
     & 24    & XDE       & XEC/XED Execute even                & Canonical \\
     & 25    & XDE       & XEC/XED Execute odd                 & Canonical \\
     & 26    & ITP       & ITP indirect cycle                  & Set to 0 \\
     & 27    & RFI       & Restart instruction                 & Canonical \\
     & 28    & ITS       & ITS indirect cycle                  & Set to 0 \\
     & 29    & FIF       & Fault during instr. fetch           & Canonical \\
     & 30-35 & CT HOLD   & Address tag hold                    & Canonical \\
6    & 0-35  & IWB       & Instruction working buffer          & Canonical \\
7    & 0-35  & IRODD     & Odd instruction                     & Canonical \\
\end{tabular}

(a) The Page Table and Segment Table associative memorys are not
implemented. They will always report disabled and no match.

(b) No parity detection is implemented, so parity error flags are never set.

(c) Operation not complete detection is not implemented, so ONC error flags are never set.

(d) Only 0 (LPRPn fault) and 8 (not control) values are returned.

(e) The TSNx values are not used by the emulator.

(f) The POT bit is used by the IC, SC and SCR address modifcation routines to mark when the
data word addressed by the indirect word is being accessed.


\part{"As-built"}

\section{Variations between the hardware and the emulator}

DPS8 memory is managed by the CPU code, not the SCU code.

The history registers are not implemented.

The RPL instruction is not implemented.

\end{document}
